# Current land use

This section examines current residential land uses in each of the three localities. The towns of Glasgow and Goshen are not included.

```{r}

library(tidyverse)
library(hdatools)
library(scales)
library(sf)
library(mapview)
library(kableExtra)

```


## Rockbridge County

```{r}

# Load Rockbridge zoning and land use data

rock_occ <- read_csv("data/rockbridge/rb-occupancy-codes.csv")
rock_landuse_zoning <- read_rds("data/rockbridge/rock_landuse_zoning.rds")
rock_area <- read_rds("data/rockbridge/rock_area.rds")

```

### Categories

Parcels in the county are assigned one of 19 unique land use codes in the real property assessment data. To standardize these specific descriptions, we can assign new summary groups and categories, as shown in the table below. Residential land uses are highlighted.

```{r}
#| label: tbl-lu-desc
#| tbl-cap: "Land use codes and descriptions assigned by county"

rock_occ |> 
  mutate(category = fct_relevel(category, "Residential", "Commercial",
                                "Agricultural", "Industrial")) |> 
  arrange(category) |> 
  select(code, description, group, category) |> 
  kable(
    col.names = c("Code", "Description", "Group", "Category"),
    align = "clll"
    ) |>
  kableExtra::row_spec(1:6, bold = TRUE, color = "#011e41") |> 
  kableExtra::kable_styling(
    bootstrap_options = c("hover", "condensed", "responsive")
    )

```

Rockbridge County includes large tracts of state and federal land, such as the George Washington National Forest and Moore's Creek State Forest. To account for these preserved areas, all parcels under public ownership were designated a new public property category by hand. The category is grouped according to which level of government owns the parcel.

The table below shows the revised land use categories and groups.

```{r}
#| label: tbl-lu-rev
#| tbl-cap: "Revised land use categories and groups"

rock_area |> 
  select(1, 2) |> 
  distinct() |> 
  mutate(category = fct_relevel(category, "Residential", "Public property",
                                "Commercial", "Agricultural", "Industrial")) |> 
  arrange(category) |> 
  kable(col.names = c("Category", "Group")) |> 
  kableExtra::kable_styling(
    bootstrap_options = c("hover", "condensed", "responsive")
    )

```

### Distribution

```{r}
#| label: rock-dist

rock_category <- rock_area |> 
  summarise(parcels = sum(parcels),
            area = sum(area),
            parcels_pct = sum(parcels_pct),
            area_pct = sum(area_pct),
            .by = c(group, category))

rock_residential <- rock_area |> 
  filter(category == "Residential") |>
  summarise(parcels = sum(parcels),
            area = sum(area),
             .by = c(group)) |> 
  mutate(parcels_pct = parcels/sum(parcels),
         area_pct = area/sum(area))

```

#### As share of parcels {.unnumbered}

There are a total of 22,777 individual parcels across Rockbridge County. Nearly all of them are either residential (46.6 percent) or vacant (45.3 percent). All other uses comprise less than 3 percent on their own.

```{r}
#| label: fig-rock-dist-parcel
#| fig-cap: "Percent of all parcels in Rockbridge County by land use category"

rock_category |> 
  summarise(parcels_pct = sum(parcels_pct), .by = c(category)) |> 
  mutate(category = fct_reorder(category, parcels_pct, .fun = sum)) |> 
  ggplot(aes(y = category)) +
  geom_col(
    aes(x = parcels_pct),
    just = 1,
    width = 0.4,
    fill = "#334a66"
    ) +
  geom_text(
    aes(
      x = parcels_pct + 0.005,
      y = category,
      label = label_percent(accuracy = 0.1)(parcels_pct)
      ),
    color = "grey50",
    hjust = 0,
    vjust = 1,
    nudge_y = -0.1
    ) +
  geom_text(
    aes(
      x = 0,
      y = category,
      label = category
      ),
    color = "#011e41",
    hjust = 0,
    vjust = 0,
    nudge_y = 0.1,
    nudge_x = 0.002
    ) +
  scale_x_continuous(
    limits = c(0, 0.49),
    expand = expansion(mult = c(0,0.05))
    ) +
  labs(
    title = "Over 9 in 10 parcels are residential or vacant",
    subtitle = "Percent of all <span style='color:#011e41;'>**parcels**</span> in Rockbridge County by land use category",
    caption = "**Source:** Rockbridge County GIS records."
  ) +
  theme_hfv() +
  add_zero_line("x") +
  theme(
    axis.text = element_blank(),
    panel.grid.major.y = element_blank()
    )

```

::: {.callout-warning}
## Interpreting the lack of "agricultural" parcels
According to county assessment records, only two parcels were designated with the "farm" description, and thus assigned to the "agricultural" category. However, we know that many more parcels in the county are used for agricultural purposes, so it is reasonable to interpret that a significant share of the "residential" and "vacant" parcels include farming activities.

For more information on agricultural parcels in the county, see analysis of the A-1 and A-2 zoning districts in [Chapter 4](existing-zoning.qmd).
:::

Within just the 10,610 residential parcels, nearly all (90.3 percent) are single-family homes. Mobile homes on individual lots make up just under 8 percent; these include single-wides and double-wides. Only 173 parcels (1.6 percent) have townhomes or condos, and 27 parcels (0.3 percent) have apartments.

```{r}
#| label: fig-rock-dist-res-parcel
#| fig-cap: "Percent of all residential parcels in Rockbridge County by land use group"

rock_residential |> 
  mutate(group = fct_reorder(group, parcels_pct)) |> 
  ggplot(aes(y = group)) +
  geom_col(
    aes(x = parcels_pct),
    just = 1,
    width = 0.6,
    fill = "#334a66"
    ) +
  geom_text(
    aes(
      x = parcels_pct + 0.005,
      y = group,
      label = label_percent(accuracy = 0.1)(parcels_pct)
      ),
    color = "grey50",
    hjust = 0,
    vjust = 1,
    nudge_y = -0.15
    ) +
  geom_text(
    aes(
      x = parcels_pct + 0.005,
      y = group,
      label = paste(number(parcels, big.mark = ","), "parcels")
      ),
    color = "grey50",
    hjust = 0,
    vjust = 1,
    nudge_y = -0.35
    ) +
  geom_text(
    aes(
      x = 0,
      y = group,
      label = group
      ),
    color = "#011e41",
    hjust = 0,
    vjust = 0,
    nudge_y = 0.1,
    nudge_x = 0.002
    ) +
  scale_x_continuous(
    limits = c(0, 0.95),
    expand = expansion(mult = c(0,0.1))
    ) +
  scale_y_discrete(
    expand = expansion(mult = c(0.1,0.1))
  ) +
  labs(
    title = "Less than 2 percent of residential parcels are multifamily",
    subtitle = "Percent of all <span style='color:#011e41;'>**residential parcels**</span> in Rockbridge County by land use group",
    caption = "**Source:** Rockbridge County GIS records."
  ) +
  theme_hfv() +
  add_zero_line("x") +
  theme(
    axis.text = element_blank(),
    panel.grid.major.y = element_blank()
    )

```

:::{.callout-tip}
## Single-family compared to multifamily parcels
For every parcel with a townhome, condo, or apartment building, there are another 48 parcels with single-family homes.
:::

#### As share of land area {.unnumbered}

Excluding Lexington and both towns, Rockbridge County is approximately 376,900 total acres. Residential and vacant parcels comprise the majority of this area, each accounting for about 35 percent of all land in the county.

Publicly-owned lands make up nearly 25 percent of the county's area, but just 1.3 percent of all parcels. These large tracts---including national forests and wildlife management areas---are permanently off limits to new development.

```{r}
#| label: fig-rock-dist-land
#| fig-cap: "Percent of all land area in Rockbridge County by land use category"

rock_category |> 
  summarise(area_pct = sum(area_pct), .by = c(category)) |> 
  mutate(category = fct_reorder(category, area_pct, .fun = sum)) |> 
  ggplot(aes(y = category)) +
  geom_col(
    aes(x = area_pct),
    just = 1,
    width = 0.4,
    fill = "#a29dd4"
    ) +
  geom_text(
    aes(
      x = area_pct + 0.005,
      y = category,
      label = label_percent(accuracy = 0.1)(area_pct)
      ),
    color = "grey50",
    hjust = 0,
    vjust = 1,
    nudge_y = -0.1
    ) +
  geom_text(
    aes(
      x = 0,
      y = category,
      label = category
      ),
    color = "#8b85ca",
    hjust = 0,
    vjust = 0,
    nudge_y = 0.1,
    nudge_x = 0.002
    ) +
  scale_x_continuous(
    limits = c(0, 0.38),
    expand = expansion(mult = c(0,0.05))
    ) +
  labs(
    title = "Large areas of the county are non-developable public land",
    subtitle = "Percent of all <span style='color:#8b85ca;'>**land area**</span> in Rockbridge County by land use category",
    caption = "**Source:** Rockbridge County GIS records."
  ) +
  theme_hfv() +
  add_zero_line("x") +
  theme(
    axis.text = element_blank(),
    panel.grid.major.y = element_blank()
    )

```

The total area of residential parcels in the county is about 132,850 acres. Single-family home properties account for nearly every acre---over 97 percent---of that land. The remaining land is almost exclusively parcels with mobile homes (2.8 percent).

The few properties with apartments, townhomes, or condos make up just over 200 acres. This is less than 0.2 percent of all the residential land in Rockbridge County.

```{r}
#| label: fig-rock-dist-res-area
#| fig-cap: "Percent of all residential land area in Rockbridge County by land use group"

rock_residential |> 
  mutate(group = fct_reorder(group, area_pct)) |> 
  ggplot(aes(y = group)) +
  geom_col(
    aes(x = area_pct),
    just = 1,
    width = 0.6,
    fill = "#a29dd4"
    ) +
  geom_text(
    aes(
      x = area_pct + 0.005,
      y = group,
      label = label_percent(accuracy = 0.1)(area_pct)
      ),
    color = "grey50",
    hjust = 0,
    vjust = 1,
    nudge_y = -0.15
    ) +
  geom_text(
    aes(
      x = area_pct + 0.005,
      y = group,
      label = paste(number(area, big.mark = ","), "acres")
      ),
    color = "grey50",
    hjust = 0,
    vjust = 1,
    nudge_y = -0.35
    ) +
  geom_text(
    aes(
      x = 0,
      y = group,
      label = group
      ),
    color = "#8b85ca",
    hjust = 0,
    vjust = 0,
    nudge_y = 0.1,
    nudge_x = 0.002
    ) +
  scale_x_continuous(
    limits = c(0, 1.1),
    expand = expansion(mult = c(0,0.1))
    ) +
  scale_y_discrete(
    expand = expansion(mult = c(0.1,0.1))
  ) +
  labs(
    title = "Almost every acre of residential land in the county is used by properties with single-family homes",
    subtitle = "Percent of all <span style='color:#8b85ca;'>**residential land area**</span> in Rockbridge County by land use group",
    caption = "**Source:** Rockbridge County GIS records."
  ) +
  theme_hfv() +
  add_zero_line("x") +
  theme(
    axis.text = element_blank(),
    panel.grid.major.y = element_blank()
    )

```

### Analysis

The chart below shows the average and median lot sizes for the four different types of residential parcels found in the county. 

```{r}
#| label: fig-rock-lot-avg
#| fig-cap: "Average and median residential lot sizes in Rockbridge County by land use group"

rock_lot_avg <- rock_landuse_zoning |> 
  st_drop_geometry() |> 
  filter(category == "Residential",
         CALCACRES > 0) |> 
  summarise(
    count = n(),
    Median = median(CALCACRES),
    Average = mean(CALCACRES),
    .by = group
    ) |> 
  pivot_longer(cols = 2:4, names_to = "lot_size")

rock_lot_avg |> 
  filter(lot_size != "count") |> 
  mutate(group = fct_reorder(group, value, .desc = FALSE)) |> 
  ggplot(aes(group, fct_rev(lot_size), color = lot_size)) +
  geom_point(
    aes(size = value),
    shape = 15
  ) + 
  geom_text(
    aes(label = paste(number(value, accuracy = 0.01), "acres")),
    vjust = 6.5
  ) +
  scale_size(range = c(2, 34)) +
  scale_color_hfv() +
  scale_y_discrete(expand = expansion(c(0.6,0.4))) +
  labs(
    title = "The average single-family home property is almost twice the size of the average apartment property",
    subtitle = "Average and median residential lot sizes in Rockbridge County by land use group",
    caption = "**Source:** Rockbridge County GIS records."
  ) +
  theme_hfv() +
  theme(
    panel.grid.major.y = element_blank(),
    axis.text.x = element_text(vjust = 5)
   )

```

At almost 14 acres, the average size of a single-family lot is about twice as large as the average apartment property (6.83 acres), despite the latter accommodating multiple homes on each lot. However, these values are right-skewed due to a small number of very large lots. Still, when using medians to account for this skew, the typical single-family lot is more than 60 percent larger than the typical lot with an apartment building.

The chart below shows the distribution of residential lot sizes by land use group.

```{r}
#| label: fig-rock-lot-avg
#| fig-cap: "Lot size distribution of residential parcels in Rockbridge County by land use group"

library(ggridges)

rock_lot_dist <- rock_landuse_zoning |> 
  st_drop_geometry() |> 
  filter(category == "Residential",
         CALCACRES > 0) |> 
  filter(
    case_when(
      group == "Single-family" ~ CALCACRES < 20,
      group == "Apartment" ~ CALCACRES < 50,
      group == "Mobile home" ~ CALCACRES < 20,
      group == "Townhouse/condo" ~ CALCACRES < 2
    )
  )

rock_lot_quantiles <- rock_landuse_zoning |> 
  st_drop_geometry() |> 
  filter(category == "Residential",
         CALCACRES > 0) |>  
  summarise(median = quantile(CALCACRES, probs = 0.5),
            upper = quantile(CALCACRES, probs = 0.75),
            .by = group)

rock_lot_dist |> 
  mutate(group = fct_reorder(group, CALCACRES, .fun = mean, .desc = TRUE)) |> 
  ggplot(aes(CALCACRES, color = group, fill = group)) +
  facet_wrap(~group, scales = "free") +
  geom_density(alpha = 0.5, linewidth = 1) +
  geom_vline(
    data = rock_lot_quantiles,
    aes(
      xintercept = upper,
      color = group
      ),
    linewidth = 1,
    alpha = 0.4,
    linetype = "dashed"
    ) +
  scale_fill_hfv() +
  scale_color_hfv() +
  scale_x_continuous(
    name = "Lot size (acres)",
    limits = c(0, NA),
    expand = c(0,0)
    ) +
  theme_hfv(flip_gridlines = TRUE) +
  add_zero_line("x") +
  theme(
    panel.spacing.x = unit(2, "lines"),
    panel.spacing.y = unit(1, "lines"),
    axis.title.x = element_text(face = "italic", vjust = -1, hjust = 0, margin = margin(t=3, b=5)),
    axis.text.y = element_blank(),
    strip.text.x = element_text(vjust = 1, hjust = 0, margin = margin(l=0, b=5))
  ) +
  labs(
    title = "Almost all residential lots are less than 2 acres in size",
    subtitle = "Lot size distribution of residential parcels in Rockbridge County by land use group",
    caption = "**Source:** Rockbridge County GIS records.<br>**Note:** Different x-axis scales are used in each panel."
  )

```


## City of Lexington

### Categories

- Descriptions of residential land use categories

### Distribution

- Residential land uses by parcel
- Residential land uses by land area

### Analysis

- Notable changes in residential land use across county over time

## City of Buena Vista

### Categories

- Descriptions of residential land use categories

### Distribution

- Residential land uses by parcel
- Residential land uses by land area

### Analysis

- Notable changes in residential land use across county over time
